---
title: "Getting started with lightgraph"
output:
  rmarkdown::html_vignette:
    fig_width: 7
vignette: >
  %\VignetteIndexEntry{Getting started with lightgraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

lightgraph is a high-performance HTML canvas-based network visualization
htmlwidget. Every figure in this vignette is **live** — drag to pan,
scroll to zoom, hover a node to light up its neighborhood, shift-drag to
box-select, and double-click a node to isolate its ego network (Escape
restores).

```{r setup}
library(lightgraph)
```

## A first graph

Edges are enough; nodes are derived automatically:

```{r first-graph}
friends <- data.frame(
  source = c("Amy", "Amy", "Ben", "Cat", "Dan", "Dan", "Eve"),
  target = c("Ben", "Cat", "Cat", "Dan", "Eve", "Fay", "Fay"),
  weight = c(3, 1, 2, 1, 2, 1, 3)
)
lightgraph(edges = friends, height = "320px")
```

A nodes data frame adds per-node attributes (`id` required; `group`,
`color`, `size` optional), and `adjacency_to_lightgraph()` converts
adjacency matrices:

```{r nodes-frame, eval = FALSE}
nodes <- data.frame(id = c("Amy", "Ben"), group = c("g1", "g2"))
lightgraph(nodes, edges)

parts <- adjacency_to_lightgraph(adj_matrix, node_names)
lightgraph(parts$nodes, parts$edges)
```

## Les Misérables

The package bundles three classic datasets: `les_mis` (weighted character
co-occurrences from Victor Hugo's novel), `got` (A Storm of Swords
character interactions), and `football_edges`/`football_nodes` (Division
IA college football games). What does the Les Misérables network look
like?

```{r summary}
str(lg_summary(les_mis))
```

`edge_weight_to_width` and `edge_weight_to_opacity` map edge weights into
stroke width and opacity, and any named numeric vector drives node size
via `node_metric` — here, weighted degree, so the protagonists stand
out:

```{r les-mis}
strength <- lg_degree(les_mis, weighted = TRUE)
lg_top_nodes(strength, 5)

lightgraph(
  edges = les_mis,
  node_metric = strength,
  metric_label = "Weighted degree",
  edge_weight_to_width = TRUE,
  weight_width_range = c(0.4, 6),
  height = "480px"
)
```

`metric_map` chooses the channel for the metric: `"size"` (default),
`"color"`, or `"both"`, interpolating between the `metric_colors`
endpoints. When the legend is on, a metric section shows the size/color
scale with its min/max values — title it with `metric_label` (e.g.
`metric_label = "PageRank"`). Here a PageRank heat map:

```{r pagerank}
lightgraph(
  edges = les_mis,
  node_metric = lg_pagerank(les_mis),
  metric_map = "both",
  metric_colors = c("#dbe9f6", "#08306b"),
  metric_label = "PageRank",
  height = "480px"
)
```

## Groups and communities

A `group` column on the nodes data frame colors nodes and builds the
interactive legend (click a legend entry to select the group). The
football network ships its ground truth — each team's conference:

```{r football}
lightgraph(
  nodes = football_nodes,
  edges = football_edges,
  show_ellipses = FALSE,
  height = "480px"
)
```

With no known grouping, `node_groups = "auto"` detects communities via
`lg_communities()` (Louvain when igraph is installed, label propagation
otherwise):

```{r got-communities}
lightgraph(
  edges = got,
  node_groups = "auto",
  node_metric = lg_pagerank(got),
  metric_label = "PageRank",
  show_ellipses = FALSE,
  height = "480px"
)
```

## Graph analytics

The `lg_*` helpers are dependency-free, treat graphs as undirected, and
return named vectors keyed by node id:

| Function | What it tells you |
|---|---|
| `lg_summary(edges)` | node/edge counts, density, degree stats, components, transitivity |
| `lg_degree(edges, weighted = FALSE)` | connections per node (weighted: strength) |
| `lg_betweenness(edges)` | who sits on the shortest paths |
| `lg_closeness(edges)` | who can reach everyone fastest |
| `lg_eigenvector(edges)` | who is connected to the well-connected |
| `lg_pagerank(edges, damping = 0.85)` | random-surfer importance |
| `lg_communities(edges)` | community labels ('c1', 'c2', ...) |
| `lg_components(edges)` | connected component ids, largest first |
| `lg_neighbors(edges, node, depth = 1)` | the k-hop ego set around a node |
| `lg_top_nodes(metric, n = 10)` | the n best entries of any metric vector |

Raw popularity and brokerage disagree — Valjean tops both, but the
betweenness ranking surfaces the story's connectors:

```{r metrics}
metrics <- data.frame(
  strength    = strength,
  betweenness = lg_betweenness(les_mis),
  pagerank    = lg_pagerank(les_mis)
)
head(metrics[order(-metrics$betweenness), ], 5)
```

`lg_components()` and `lg_neighbors()` slice a graph:

```{r slicing}
reach <- lg_neighbors(les_mis, "Fantine", depth = 1)
sub <- les_mis[les_mis$source %in% reach & les_mis$target %in% reach, ]
nrow(sub)
```

## Interaction and themes

Everything is modeless — no tool switching. `ego_depth` sets how many
hops a double-click reveals, `neighbor_fade` controls how hard
non-neighbors fade on hover, and `theme = "dark"` swaps canvas, panels,
labels, and default colors:

```{r dark}
lightgraph(
  edges = got,
  node_groups = "auto",
  theme = "dark",
  ego_depth = 2,
  neighbor_fade = 0.06,
  show_ellipses = FALSE,
  height = "480px"
)
```

## The config escape hatch

Arguments cover the common surface; `config` deep-merges a raw lightGraph
configuration (the JS shape) over the generated one, so any option in the
JS `DEFAULT_CONFIG` is reachable:

```{r config, eval = FALSE}
lightgraph(
  edges = les_mis,
  config = list(
    nodes = list(borderWidth = 2, selectedBorderColor = "#ffd700"),
    labels = list(fontFamily = "Georgia, serif"),
    simulation = list(centerStrength = 0.8)
  )
)
```

## Saving and Shiny

```{r saving, eval = FALSE}
w <- lightgraph(edges = les_mis)

htmlwidgets::saveWidget(w, "les_mis.html")  # standalone file

# Shiny
ui <- fluidPage(lightgraphOutput("net"))
server <- function(input, output) {
  output$net <- renderLightgraph(lightgraph(edges = les_mis))
}
```

Viewers can also export PNG, SVG, and JSON from the widget toolbar.

## Learn more

The full feature tour — including a 3,257-airport world flight network,
metric-driven styling, circular layouts, and performance tuning — lives
in the [online documentation](https://haozhu233.github.io/lightgraph/),
and every argument has the same name and meaning in the Python package.
